### 2020-03-02 doc/icfp20/intro.tex
comment out some text

Full dependent types allow expression of predicate logic, and it is
tempting to take a standard textbook on music theory such as
\citet{piston-harmony} or \citet{aldwell2018harmony} and 
formalize it in type theory. However this is difficult because most
``rules'' in music theory are not absolute, but rather suggestions of
varying degrees (themselves vaguely specified) of importance. Since
ultimately the resulting created music is more important than the
theory, an alternative is to attempt to indirectly deduce the rules
from music samples, using machine learning~\citep{huang-cp} or
some other statistical methods. Unfortunately the set of deduced
rules could sometimes be insufficient for generating good music.
As an example, \citet{roberts-cp} develop the Bach Doodle
harmonization tool using Bach chorals as training data, but still
observe high frequency of parallel 5th and 8th that are extremely
rare in Bach's music.

Ultimately it seems some kind of mixture of rules and statistics may
be the best way to faithfully capture an effective theory of
music. One simple approach would be to develop a logic of rules and
their interactions, but attach weights to the rules which could be
determined via machine learning techniques.

Our long-term goal is formalization of a large amount of music theory.
As a first step we start with the modest task of expressing a small,
relatively strict rule set known as species
counterpoint~\cite{fux-cp}, intended for combining interdependent
melody lines to produce a pleasant-sounding result.

We make use of and extend Music Tools~\citep{MusicTools},
a library of small tools
for analysis and synthesis of music written in Agda, with the goal of
eventually being a dependently-typed alternative to Euterpea.
Previous work by \citet{cong-cp} expresses
first-species counterpoint with the Agda type system, so that an error
in counterpoint resulted in an Agda type error. Although this takes
advantage of Agda's native type checking, it has several
downsides. One is that the Agda errors may be difficult to interpret
as the corresponding musical error. Another is that it is difficult to
keep the types both simple and flexible.

In this work we present an alternative representation of counterpoint
using what could be considered a custom musical type checker written
in Agda. This allows us to easily describe fine-grained rules, write
custom type errors, and then combine subsets of rules together for
larger-scale checking. Music can be easily combined with rules it
satisfies as a dependent record type. This record represents a proof
certificate that the music follows the asserted rules.
Rules can also be used in multiple different contexts, satisfying
modularity and composability.

We also present preliminary work on harmonizing a melody using
a subset of rules based on \citet{piston-harmony}, contrasting with
existing work by \citet{koops-fharm}. Notably, since counterpoint and
harmony are not separate concepts but in fact deeply intertwined, we
are able to reuse counterpoint rules to help develop natural-sounding
harmonizations.

This is an experience report, and we highlight both the advantages and
disadvantages of using Agda for music theory. On one hand the
expressiveness of dependent types makes it easy and natural to
describe music theory rules. On the other hand we find the emphasis on proof
construction and particularly the extra work needed for decidable
equality can add an extra burden which is not always welcome. However
overall we feel the positives far outweigh the negatives, and in the
final section we describe future work which may help reduce the tedium.

>>>>

[NOTE: I've commented out the discussion on explicit / implicit
handling of rules, to make the flow smoother.]

% Full dependent types allow expression of predicate logic, and it is
% tempting to take a standard textbook on music theory such as
% \citet{piston-harmony} or \citet{aldwell2018harmony} and 
% formalize it in type theory. However this is difficult because most
% ``rules'' in music theory are not absolute, but rather suggestions of
% varying degrees (themselves vaguely specified) of importance. Since
% ultimately the resulting created music is more important than the
% theory, an alternative is to attempt to indirectly deduce the rules
% from music samples, using machine learning~\citep{huang-cp} or
% some other statistical methods. Unfortunately the set of deduced
% rules could sometimes be insufficient for generating good music.
% As an example, \citet{roberts-cp} develop the Bach Doodle
% harmonization tool using Bach chorals as training data, but still
% observe high frequency of parallel 5th and 8th that are extremely
% rare in Bach's music.

% Ultimately it seems some kind of mixture of rules and statistics may
% be the best way to faithfully capture an effective theory of
% music. One simple approach would be to develop a logic of rules and
% their interactions, but attach weights to the rules which could be
% determined via machine learning techniques.

Our long-term goal is to formalize a large amount of music theory.
As a first step we start with the modest task of expressing a small,
relatively strict rule set known as species
counterpoint~\cite{fux-cp}, intended for combining interdependent
melody lines to produce a pleasant-sounding result.
Roughly speaking, what we do is to write a custom musical type
checker in Agda.
Compared to the previous work by \citet{cong-cp}, where rules are
expressed directly as Agda types, our approach makes it easier to
describe fine-grained rules, produce readable error messages, and
add or drop rules depending on the circumstances.

[NOTE: I've squeezed the description of our new counterpoint
implementation.]

% We make use of and extend Music Tools~\citep{MusicTools},
% a library of small tools
% for analysis and synthesis of music written in Agda, with the goal of
% eventually being a dependently-typed alternative to Euterpea.
% Previous work by \citet{cong-cp} expresses
% first-species counterpoint with the Agda type system, so that an error
% in counterpoint results in an Agda type error. Although this takes
% advantage of Agda's native type checking, it has several
% downsides. One is that the Agda errors may be difficult to interpret
% as the corresponding musical error. Another is that it is difficult to
% keep the types both simple and flexible.

% In this work we present an alternative representation of counterpoint
% using what could be considered a custom musical type checker written
% in Agda. This allows us to easily describe fine-grained rules, write
% custom type errors, and then combine subsets of rules together for
% larger-scale checking. Music can be combined with rules it
% satisfies as a dependent record type. This record represents a proof
% certificate that the music follows the asserted rules.
% Rules can also be used in multiple different contexts, satisfying
% modularity and composability.

We also present preliminary work on harmonizing a melody using
a subset of rules based on \citet{piston-harmony}, contrasting with
existing work by \citet{koops-fharm}. Since counterpoint and
harmony are not separate concepts but in fact deeply intertwined, we
would wish to reuse counterpoint rules to develop natural-sounding
harmonizations.  We show that, with our custom type checker, it is
straightforward to reuse rules in different contexts.  Put differently,
our representation of rules satisfies modularity.

The rest of this paper structured as follows.  Section~\ref{sec:music}
provides a brief glossary of musical terms, to prepare the
reader with no familiarity in music theory.  Sections~\ref{sec:cp} and
\ref{sec:harmony} describe our formalization of counterpoint and
harmony, focusing on the composable and modular aspects of the
proposed approach.  Lastly, Section~\ref{sec:conclusion} concludes
the paper with future perspectives.

This is an experience report, and we highlight both the advantages and
disadvantages of using Agda for music theory. On one hand the
expressiveness of dependent types makes it easy and natural to
describe music theory rules. On the other hand we find the emphasis on proof
construction and particularly the extra work needed for decidable
equality can add an extra burden which is not always welcome. However
overall we feel the positives far outweigh the negatives, and in the
final section we discuss possible ways to reduce the tedium.
